---
title: "SPiCT"
author: "Martin W. Pedersen"
date: '`16-08-2016`'
output:
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
vignette: |
  %\VignetteIndexEntry{SPiCT} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = 'center')
```

This vignette explains basic and more advanced functions of the `spict` package. When loading the package you are notified which version of the package you have installed:

```{r}
library(spict)
```
The printed version follows the format ver@SHA, where ver is the manually defined version number and SHA refers to a unique commit on [github](http://github.com/mawp/spict). The content of this vignette pertains to the version printed above. 

# Loading built-in example data 

The package contains the catch and index data analysed in Polacheck et al. (1993). This data can be loaded and plotted by typing

```{r}
data(pol) 
```

Data on three stocks are contained in this dataset. Here focus will be on the South Atlantic albacore data. This dataset contains the following

```{r}
pol$albacore
```

Note that data are structure in a list containing the entries `obsC` (catch observations), `timeC` (time of catch observations), `obsI` (index observations), and `timeI` (time of index observations). If times are not specified it is assumed that the first observation is observed at time 1 and then sequentially onward with a time step of one year. It is therefore recommended to always specify observation times.

# Plotting data

The data can be plotted using the command

```{r, fig.width=5, fig.height=5.5, out.width='0.5\\textwidth', fig.show='hold'}
plotspict.data(pol$albacore)
```

Note that the number of catch and index observations are given in the respective plot headers. Furthermore, note colour legend in the top-right corner, which explains at what time of the year an observation was observed based on the colour of the individual points. For illustrative purposes let's try shifting the data a bit

```{r, fig.width=5, fig.height=5.5, out.width='0.5\\textwidth', fig.show='hold'}
inpshift <- pol$albacore
inpshift$timeC <- inpshift$timeC + 0.3
inpshift$timeI <- inpshift$timeI + 0.8
plotspict.data(inpshift)
```

Now the colours show that catches are observed in the spring and index in the autumn.

# Advanced data plotting

There is also a more advanced function for plotting data, which at the same time does some basic model fitting (linear regression) and shows the results

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
plotspict.ci(pol$albacore)
```

The two top plots come from `plotspict.data`, with the dashed horizontal line representing a guess of MSY. This guess comes from a linear regression between the index and the catch divided by the index (middle row, left). This regression is expected to have a negative slope. A similar plot can be made showing catch versus catch/index (middle row, right) to approximatively find the optimal effort (or effort proxy). The proportional increase in the index as a function of catch (bottom row, right) should show primarily positive increases in index at low catches and vice versa. Positive increases in index at large catches could indicate model violations. In the current plot these are not seen.

# Fitting the model

The model is fitted to data by running

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
inp <- pol$albacore
inp$timepredc <- 1993 # Prediction horizon for catches
system.time(rep <- fit.spict(inp))
```

Here the call to `fit.spict` is wrapped in the `system.time` command to check the time spent on the calculations. This is obviously not required, but done here to show that fitting the model only takes a few seconds. The result of the model fit is stored in `rep`, which can either be plotted using `plot` or summarised using `summary`. 

The results are returned as a list that contains output as well as input. The content of this list is

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
names(rep)
```

Many of these variables are generated by `TMB::sdreport()`. In addition to these `spict` includes the list of input values (`inp`), the object used for fitting (`obj`), the result from the optimiser (`opt`), the time spent on fitting the model (`computing.time`), and more less useful variables.

# Interpreting the summary of results

# Interpreting the plot of results

The results are plotted using

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
plot(rep)
```

Some general comments can be made regarding the style and colours of these plots:

- Estimates (biomass, fishing mortality, catch, production) are shown using blue lines.
- 95% CIs of absolute quantities are shown using dashed blue lines.
- 95% CIs of relative biomass and fishing mortality are shown using shaded blue regions.
- Estimates of reference points ($B_{MSY}$, $F_{MSY}$, $MSY$) are shown using black lines.
- 95% CIs of reference points are shown using grey shaded regions.
- The end of the data range is shown using a vertical grey line.
- Predictions beyond the data range are shown using dotted blue lines.
- Data are shown using points coloured by season. Different index series use different point characters (not shown here).

The individual plots can be plotted separately. We will now look at them one at a time. The top left is the plot of absolute biomass

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
plotspict.biomass(rep)
```

Note that this plot has a y-axis on the right side related to the relative biomass ($B_t/B_{MSY}$). The shaded 95% CI region relates to this axis, while the dashed blue lines relate to the left y-axis indicating absolute levels. The dashed lines and the shaded region are shown on the same plot to make it easier to assess whether the relative or absolute levels are most accurately estimated. Here, the absolute are more accurate than the relative. Later, we will see examples of the opposite. The horizontal black line is the estimate of $B_{MSY}$ with 95% CI shown as a grey region.

The plot of the relative biomass is produced using

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
plotspict.bbmsy(rep)
```

This plot contains much of the same information as given by `plotspict.biomass`, but without the information about absolute biomass and without the 95% CI around the $B_{MSY}$ reference point.

The plots of fishing mortality follow the same principles

```{r, results='show', message=FALSE, warning=FALSE, fig.width=3, fig.height=3.3, fig.show='hold'}
plotspict.f(rep, main='', qlegend=FALSE, rel.axes=FALSE, rel.ci=FALSE)
plotspict.ffmsy(rep, main='', qlegend=FALSE)
```

The estimate of $F_{MSY}$ is shown with a horizontal black line with 95% CI shown as a grey region (left plot). The 95% CI of $F_{MSY}$ is very wide in this case. As shown here it is quite straightforward to remove the information about relative levels from the plot of absolute fishing mortality. Furthermore, the argument `main=''` removes the heading and `qlegend=FALSE` removes the colour legend for data points.

The plot of the catch is produced using

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
plotspict.catch(rep)
```

This plot shows estimated catches (blue line) versus observed catches (points) with the estimate of $MSY$ plotted as a horizontal black line with its 95% CI given by the grey region.

# Extracting estimates

To extract an estimated quantity, here `logBmsy` use

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
get.par('logBmsy', rep)
```

This returns a vector with `ll` being the lower 95% limit of the CI, `est` being the estimated value, `ul` being the upper 95% limit of the CI, `sd` being the standard deviation of the estimate, and `cv` being the coefficient of variation of the estimate. The estimated quantity can also be returned on the natural scale (as opposed to log scale) by running

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
get.par('logBmsy', rep, exp=TRUE)
```

This essentially takes the exponential of `ll`, `est` and `ul` of the values in log, while `sd` is unchanged as it is the standard deviation of the quantity on the scale that it is estimated (here log). When transforming using `exp=TRUE` the $CV = \sqrt{e^{\sigma^2}-1}$. Most parameters are log-transformed under estimation and should therefore be extracted using `exp=TRUE´.

For a standard fit (not using robust observation error, seasonality etc.), the quantities that can be extracted using this method are

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
sort(unique(c(names(rep$value), names(rep$par.fixed), names(rep$par.random))))
```

These should be relatively self-explanatory when knowing that reference points ending with `s` are stochastic and those ending with `d` are deterministic, quantities ending with `p` are predictions and quantities ending with `l` are estimates in the final year. If a quantity is available both on natural and log scale and it is preferred to transform the quantity from log as most quantities are estimated on the log scale.

# Setting initial parameter values

Initial parameter values used as starting guess of the optimiser can be set using `inp$ini`. For example, to specify the initial value of `logK` set

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
inp$ini$logK <- log(100)
```

This procedure generalises to all other model parameters. If initial values are not specified they are set to default values. To see the default initial value of a parameter, here `logK`, run

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
inp <- check.inp(pol$albacore)
inp$ini$logK
```

This can also be done posterior to fitting the model by printing `rep$inp$ini$logK`.

## Checking robustness to initial parameter values

# Phases and how to fix parameters

The `spict`-package has the ability to estimate parameters in phases. Users familiar with AD model builder will know that this means that some parameters are held constant in phase 1, some are then released and estimated in phase 2, more are released in phase 3 etc. until all parameters are estimated. Per default all parameters are estimated in phase 1. As an example the standard deviation on the biomass process, `logsdb`, is estimated in phase 2:

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
inp <- pol$albacore
inp$phases$logsdb <- 2
rep <- fit.spict(inp)
```

Phases can also be used to fix parameters to their initial value by setting the phase to `-1`. For example

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
inp <- pol$albacore
inp$phases$logsdb <- -1
inp$ini$logsdb <- log(0.1)
rep <- fit.spict(inp)
summary(rep)
```


# Priors

# Seasonality

# Robust estimation (reducing influence of outliers)

Estimation becomes less stable

Perhaps fix some parameters

# Simulation

# Forecasting

To make a catch forecast a forecast interval needs to be specified. This is done by specifying the start of the interval (`inp$timepredc`) and the length of the interval in years (`inp$dtpredc`). In addition to the forecast interval a fishing scenario needs to be specified. This is done by specifying a factor (`inp$ffac`) to multiply the current fishing mortality by (i.e. the F at the last time point of the time period where data are available) and the time that management should start (`inp$manstart`). The time point of the reported forecast of biomass and fishing mortality can be controlled by setting `inp$timepredi`. Producing short-term forecasts entails minimal additional computing time.

Forecasts are produced as part of the usual model fitting. To illustrate the procedure, a short example using the South Atlantic albacore dataset of Polacheck et al. (1993) containing catch and commercial CPUE data in the interval 1967 to 1989 is presented. The code to obtain the forecasted annual catch in the interval starting 1991 under a management scenario where the fishing pressure is reduced by 25% starting in 1991, and a forecasted index in 1992 is:

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
library(spict)
data(pol)
inp <- pol$albacore
inp$manstart <- 1991
inp$timepredc <- 1991
inp$dtpredc <- 1
inp$timepredi <- 1992
inp$ffac <- 0.75
rep <- fit.spict(inp)
```

To specifically show forecast results use

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
sumspict.predictions(rep)
```

This output is also shown when using `summary(rep)`. The results can be plotted using `plot(rep)`, however to visualise the change in forecasted fishing mortality and associated change in forecasted catch more clearly we use

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
par(mfrow=c(2, 2), mar=c(4, 4.5, 3, 3.5))
plotspict.bbmsy(rep, stamp='')
plotspict.ffmsy(rep, qlegend=FALSE, stamp='')
plotspict.catch(rep, qlegend=FALSE, stamp='')
plotspict.fb(rep, man.legend=FALSE, stamp='')
```

Note in the plot that the decrease in fishing pressure results in a constant biomass as opposed to the expected decrease if fishing effort had remained constant.

SPiCT has a function that runs several predefined management scenarios, which can be presented in a forecast table. To perform the calculations required to produce the forecast table run:

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
rep <- manage(rep)
```

where `rep` is the result of `fit.spict()` from the code above. Then, the results can be summarised (and extracted) by running:

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
df <- mansummary(rep)
```

Then, `df` is a data frame with each line containing a line of the output

```{r, results='show', message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.show='hold'}
head(df)
```

The resulting biomass, fishing mortality and catch of the management scenarios are included in the standard plots

```{r, results='show', message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5.5, fig.show='hold'}
par(mfrow=c(2, 2), mar=c(4, 4.5, 3, 3.5))
plotspict.bbmsy(rep, stamp='')
plotspict.ffmsy(rep, qlegend=FALSE, stamp='')
plotspict.catch(rep, qlegend=FALSE, stamp='')
plotspict.fb(rep, man.legend=FALSE, stamp='')
```




## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
